---
title: 'Reproducible Research: Peer Assessment 1'
output:
html_document:
keep_md: yes
pdf_document: default
author: "Didac Busquets"
---


## Loading and preprocessing the data

Load the CSV data and convert the 'date' column to Date format:

```{r}
activity <- read.csv('activity.csv', sep = ',')
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?

Sum the number of steps for each date, ignoring the missing values:

```{r}
total_steps_per_day <- tapply(activity$steps,activity$date, sum, na.rm=TRUE)
```

Plot the histogram of total number of steps per day:
```{r}
hist(total_steps_per_day, xlab = "Total steps per day", 
     main = "Histogram of total steps per day")
```

Compute mean and median of total number of steps per day:

```{r}
mean(total_steps_per_day)
median(total_steps_per_day)
```

## What is the average daily activity pattern?

Convert the intervals from numeric value to time (e.g. 235 to a time 02:35) and add it as a new column:

```{r}
activity$time_interval <- strptime(sprintf("%04d", as.numeric(activity$interval)), 
                      format="%H%M")
```

Compute the mean number of steps for each interval, ignoring missing values:

```{r}
mean_steps_per_interval <- tapply(activity$steps,activity$interval, 
                                  mean, na.rm=TRUE)
```

Now plot the mean number of steps per interval:

```{r}
plot(activity$time_interval[1:288], mean_steps_per_interval, type='l', 
     xlab='Interval', ylab='Mean number of steps', 
     main = "Daily activity pattern (averaged)")
```

Search for the interval with the maximum number of steps:

```{r}
max_interval <- which.max(mean_steps_per_interval)
mean_steps_per_interval[max_interval]
```

The interval with maximum number of steps is `r names(max_interval)`, with `r mean_steps_per_interval[max_interval]` steps.


## Imputing missing values

Look for missing values (NA) and calculate how many there are:

```{r}
nas <- activity$steps[is.na(activity)]
length(nas)
```

There are `r length(nas)` intervals with missing values.

The missing values will be replaced by the mean value of the given 5-minute interval. First replicate the vector with mean number of steps per interval, so that it has the same lenght as the activity data (61 days):

```{r}
repeated_means <- rep(mean_steps_per_interval, times=61)
```

Now copy the activity to a new dataset and assign these values only to missing entries:

```{r}
new_activity <- activity
new_activity[is.na(activity)] <- repeated_means[is.na(activity)]
```

Sum the number of steps for each date in the new dataset:

```{r}
total_steps_per_day_new <- tapply(new_activity$steps,new_activity$date, sum)
```

Plot the histogram of total number of steps per day:
```{r}
hist(total_steps_per_day_new, xlab = "Total steps per day", main = "Histogram of total steps per day")
```

Compute mean and median of new dataset:

```{r}
mean(total_steps_per_day_new)
median(total_steps_per_day_new)
```

These values differ from the mean and median when missing values were ignored. **This shows that one has to be careful when replacing missing values!** 

## Are there differences in activity patterns between weekdays and weekends?

Create the factor variable indicating whether a given day is weekday or weekend:

```{r}
new_activity$day_type <- factor(weekdays(new_activity$date) %in% c("Saturday","Sunday"),
                                labels = c("weekday", "weekend"))
```

Compute the mean number of steps for each interval, for weekdays and weekends:

```{r}
mean_steps_per_interval_weekday <- tapply(new_activity$steps[new_activity$day_type=='weekday'],
                                          new_activity$interval[new_activity$day_type=='weekday'],
                                          mean)

mean_steps_per_interval_weekend <- tapply(new_activity$steps[new_activity$day_type=='weekend'],
                                          new_activity$interval[new_activity$day_type=='weekend'], 
                                          mean)
```

Create two data frames containing each interval, its mean number of steps, and the type of day (weekday or weekend) and combine them (rbind) into a new single data frame:

```{r}
df_weekdays <- data.frame("interval" = new_activity$time_interval[1:288], 
                          "mean_steps" = mean_steps_per_interval_weekday, 
                          "day_type" = "weekday")

df_weekends <- data.frame("interval" = new_activity$time_interval[1:288],
                          "mean_steps" = mean_steps_per_interval_weekend, 
                          "day_type" = "weekend")

df_mean_steps <- rbind(df_weekdays,df_weekends)
```

Plot the time series of average number of steps per interval, split by day type:

```{r}
library(lattice)
xyplot(mean_steps ~ interval | day_type, data = df_mean_steps, 
       type="l", layout=c(1,2), xlab = "Interval", ylab = "Number of steps",
       main = "Daily activity pattern (averaged)",
       scales=list(
     x=list(at= seq(as.POSIXct("2015-03-10"), by="3 hour", length=24), 
     labels=format(seq(as.POSIXct("2015-03-10"), by="3 hour", length=24),
                   "%H:%M"))))
```